The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-10-11 conf_article/neon.tex
continued with updates in introductiuon

That is what can be said from a practical point of view. \replaced[id=IK, comment={Rewrote this to be more precise. The innovation wasn't just 'using' a norm, but deriving the update from a principled, constrained optimization problem. I think this better reflects the contribution of the Muon paper.}]{From a theoretical perspective, a key innovation of Muon was its principled derivation of the update rule, which emerges as the solution to an optimization problem constrained by the spectral norm \citep{bernstein2025deriving}}{From the perspective of theory, the main innovation of Muon has been an intentional usage of matrix norms, i.e. the spectral norm, to derive the algorithm's update}. Based upon recent theoretical advances that explain some theory behind Muon, Scion and Gluon \citep{bernstein2025deriving,kovalev2025understanding,pethick2025training,riabinin2025gluon}, we explore application of other matrix norms to optimization of functions of matrices. As it has been done with Muon, we stipulate that our algorithms' updates be fast to compute.

In this article, \replaced[id=IK, comment={I think this is much clearer to introduce all algorithms like that, the paragraphs where one of every two words is a ***on are confusing.}]{we generalize the spectral-norm-based approach by first leveraging the family of Ky Fan k-norms to derive a new class of algorithms, which we call Fanions. Subsequently, we create a second, hybrid family named F-Fanions, by constructing convex combinations of the Ky Fan norms with the Frobenius norm and taking dual of that composite norm. Working within the linear minimization oracle (lmo) framework~--- which is equivalent (up to a factor) to the trust region approach~--- we derive the explicit update formulas for both algorithm families. Later in \cref{sec:matrix-side} we explain how their updates can be approximated with the same $\cO(m n \min\{m,n\})$ complexity as Muon updates are computed via Newton-Schulz iterations.}{we focus on the two most common matrix norms akin to the spectral norm, namely the nuclear norm and the Frobenius norm. Working in the linear minimization oracle (lmo) framework, which is equivalent to a factor to the trust region approach and the steepest descent under norm constraint, we derive Neon, our algorithm based on the nuclear norm. In the section {\it Matrix side of the updates}, we explain how Neon updates can be computed asymptotically with the same $\cO(m n \min\{m, n\})$ number of FLOPs as Muon updates are computed via Newton-Schulz iterations.}

Noticing that Neon and Muon are diametrical in terms of the rank of the update matrix, we bridge the space by considering the dual to the Ky Fan $k$-norm, the primal norm being the sum of the $k$ largest singular values of the matrix. The resulting update turns out to be Dion \citet{ahn2025dioncommunicationefficientoptimizerlarge} without error feedback.

To make Neon and Dion without error feedback more robust, we "regularize" them by Normalized SGD (NSGD), which is derived in lmo with the Frobenius norm. We do this in the same lmo approach by considering a norm that is dual to the convex combination of the Frobenius norm and the Ky Fan k-norm. So we derive the algorithms we name F-Neon and F-Dion respectively. The same procedure we follow with Muon to obtain F-Muon.

Having faced the array of different Muon-like optimizers, which, according to the upper bounds from \citet{kovalev2025understanding,riabinin2025gluon}, must converge similarly, we painstakingly compare them on a synthetic linear least squares problem with a known Lipschitz constant. The efforts result in comparison of the algorithms by their convergence not in terms of the dual norms of the gradient, but in terms of the common spectral norm, which may strongly differ, especially in large matrices, from the initial dual norms. Thus, we compare the algorithms in a unified fashion and find that sometimes if an algorithm converges fast by the loss, it does not imply its fast convergence by the norm of the gradient, and vice versa.

Finally, observing the poor performace of Neon and Dion without error feedback, which negatively impacts F-Neon and F-Dion as well, we test Muon and F-Muon on two deep-learning tasks, training convolutional network on CIFAR-10 \citep{cifar2023airbench} and pre-training NanoGPT on FineWeb \citep{modded_nanogpt_2024}. The results support the supremacy of Muon, but the most striking outcome of the tests is that F-Muon, only half of which is Muon, is on par with Muon on training the convolutional network and only slighter worse that Muon on NanoGPT pre-training, while NSGD is far from that. The case of F-Muon answers in the affirmative to the question of possibility to perturb the norm that underpins the algorithm without sacrificing its performace. We envision that this effect will be used in the future to constuct easier to compute approximations of empirically optimal algorithms like Muon.

>>>>

That is what can be said from a practical point of view. \replaced[id=IK, comment={Rewrote this to be more precise. The innovation wasn't just 'using' a norm, but deriving the update from a principled, constrained optimization problem. I think this better reflects the contribution of the Muon paper.}]{From a theoretical perspective, a key innovation of Muon was its principled derivation of the update rule, which emerges as the solution to an optimization problem constrained by the spectral norm \citep{bernstein2025deriving}}{From the perspective of theory, the main innovation of Muon has been an intentional usage of matrix norms, i.e. the spectral norm, to derive the algorithm's update}.

Motivated by the success of Muon, many generalizations and variations of it were proposed. Among the notable ones are Scion \citep{}, Deon \citep{} and Gluon \citep{}. One question still persists however: Why

Based upon recent theoretical advances that explain some theory behind Muon, Scion and Gluon \citep{bernstein2025deriving,kovalev2025understanding,pethick2025training,riabinin2025gluon}, we explore application of other matrix norms to optimization of functions of matrices. As it has been done with Muon, we stipulate that our algorithms' updates be fast to compute.

\replaced[id = IK, comment={Subsequent paragraphs in the introduction are almost completely new.}]{}{}

%In this article, \replaced[id=IK, comment={I think this is much clearer to introduce all algorithms like that, the paragraphs where one of every two words is a ***on are confusing.}]{we generalize the spectral-norm-based approach by first leveraging the family of Ky Fan k-norms to derive a new class of algorithms, which we call Fanions. Subsequently, we create a second, hybrid family named F-Fanions, by constructing convex combinations of the Ky Fan norms with the Frobenius norm and taking dual of that composite norm. Working within the linear minimization oracle (lmo) framework~--- which is equivalent (up to a factor) to the trust region approach~--- we derive the explicit update formulas for both algorithm families. Later in \cref{sec:matrix-side} we explain how their updates can be approximated with the same $\cO(m n \min\{m,n\})$ complexity as Muon updates are computed via Newton-Schulz iterations.}{we focus on the two most common matrix norms akin to the spectral norm, namely the nuclear norm and the Frobenius norm. Working in the linear minimization oracle (lmo) framework, which is equivalent to a factor to the trust region approach and the steepest descent under norm constraint, we derive Neon, our algorithm based on the nuclear norm. In the section {\it Matrix side of the updates}, we explain how Neon updates can be computed asymptotically with the same $\cO(m n \min\{m, n\})$ number of FLOPs as Muon updates are computed via Newton-Schulz iterations.}

%Noticing that Neon and Muon are diametrical in terms of the rank of the update matrix, we bridge the space by considering the dual to the Ky Fan $k$-norm, the primal norm being the sum of the $k$ largest singular values of the matrix. The resulting update turns out to be Dion \citet{ahn2025dioncommunicationefficientoptimizerlarge} without error feedback.

%To make Neon and Dion without error feedback more robust, we "regularize" them by Normalized SGD (NSGD), which is derived in lmo with the Frobenius norm. We do this in the same lmo approach by considering a norm that is dual to the convex combination of the Frobenius norm and the Ky Fan k-norm. So we derive the algorithms we name F-Neon and F-Dion respectively. The same procedure we follow with Muon to obtain F-Muon.

%Having faced the array of different Muon-like optimizers, which, according to the upper bounds from \citet{kovalev2025understanding,riabinin2025gluon}, must converge similarly, we painstakingly compare them on a synthetic linear least squares problem with a known Lipschitz constant. The efforts result in comparison of the algorithms by their convergence not in terms of the dual norms of the gradient, but in terms of the common spectral norm, which may strongly differ, especially in large matrices, from the initial dual norms. Thus, we compare the algorithms in a unified fashion and find that sometimes if an algorithm converges fast by the loss, it does not imply its fast convergence by the norm of the gradient, and vice versa.

%Finally, observing the poor performace of Neon and Dion without error feedback, which negatively impacts F-Neon and F-Dion as well, we test Muon and F-Muon on two deep-learning tasks, training convolutional network on CIFAR-10 \citep{cifar2023airbench} and pre-training NanoGPT on FineWeb \citep{modded_nanogpt_2024}. The results support the supremacy of Muon, but the most striking outcome of the tests is that F-Muon, only half of which is Muon, is on par with Muon on training the convolutional network and only slighter worse that Muon on NanoGPT pre-training, while NSGD is far from that. The case of F-Muon answers in the affirmative to the question of possibility to perturb the norm that underpins the algorithm without sacrificing its performace. We envision that this effect will be used in the future to constuct easier to compute approximations of empirically optimal algorithms like Muon.
